---
title: "CMHtest 2x2 idea: an LRstats/loglinear analog"
author: "Michael Friendly"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, message = FALSE, warning = FALSE)
```

See `dev/CMH-2x2.md` for the design notes this supports. That file sketches reorganizing
`CMHtest()`'s four statistics (`cor`, `rmeans`, `cmeans`, `general`) into a 2x2 layout crossing
how the row/column variables are each treated -- general (nominal) vs. ordered (scored). This Rmd
checks a loglinear-model analog of the same idea, using the model-comparison example already in
`Mental`'s own `@examples` (`R/data.R`, ~lines 2478-2498).

```{r load}
library(vcdExtra)
data(Mental)
```

## The CMH side

```{r cmh}
cmh <- CMHtest(Freq ~ ses + mental, data = Mental)
print(cmh)
```

## The loglinear-model side

Four association terms, each treating `ses` (row, 6 levels) and `mental` (column, 4 levels) as
either its own factor (general/nominal) or via a single integer score (ordered):

```{r models}
indep <- glm(Freq ~ mental + ses, family = poisson, data = Mental)

Cscore <- as.numeric(Mental$ses)     # a "row" score, despite the name
Rscore <- as.numeric(Mental$mental)  # a "column" score, despite the name

linlin <- update(indep, . ~ . + Rscore:Cscore)  # both ordered
roweff <- update(indep, . ~ . + mental:Cscore)  # mental (col) general, ses (row) ordered
coleff <- update(indep, . ~ . + Rscore:ses)     # ses (row) general, mental (col) ordered
rowcol <- update(indep, . ~ . + Rscore:ses + mental:Cscore)  # both terms together

LRstats(indep, linlin, roweff, coleff, rowcol)
```

**Note the naming trap**: `roweff` uses `mental:Cscore` -- i.e. `mental` (the *column* variable)
keeps its own factor effects, while `ses` (row) is reduced to the single score `Cscore`. That
structurally matches CMH's `cmeans` ("col mean scores differ" -- row ordered, col general), not
`rmeans`, despite the name. `coleff` is the mirror image, matching `rmeans`. The variable names
describe which *score* is used, not which CMH cell the model lands in.

## Correspondence check

LR tests of each model against `indep`, compared to the matching `CMHtest()` row:

```{r compare}
lr_test <- function(mod, label) {
  a <- anova(indep, mod, test = "Chisq")
  data.frame(model = label, LRchisq = round(a$Deviance[2], 3), df = a$Df[2],
             p = signif(a[["Pr(>Chi)"]][2], 4))
}

saturated <- glm(Freq ~ mental * ses, family = poisson, data = Mental)

lr_results <- rbind(
  lr_test(linlin, "linlin"),
  lr_test(roweff, "roweff"),
  lr_test(coleff, "coleff"),
  lr_test(saturated, "saturated (mental*ses)")
)

cmh_results <- data.frame(
  cmh_type = c("cor", "cmeans", "rmeans", "general"),
  cmh_chisq = cmh$table[c("cor", "cmeans", "rmeans", "general"), "Chisq"],
  cmh_df = cmh$table[c("cor", "cmeans", "rmeans", "general"), "Df"]
)

cbind(cmh_results, lr_results[, -1])
```

Each pair lines up closely (same df, similar magnitude) -- expected, since CMH and LR tests are
asymptotically equivalent ways of testing the same association hypotheses. `general` needs the
**fully saturated** model, not `rowcol`.

## `rowcol` is a different, more restrictive model -- not the analog of `general`

```{r rowcol-df}
anova(indep, rowcol, test = "Chisq")
```

Both interaction terms together only reach df=7, not the df=15 of full general association --
there's a 1-df redundancy between the two term sets when combined. `rowcol` is a genuinely
different ("additive row + column effects") model, not a shortcut to the saturated one.

## A rigorous alternative to the CMH "corner" cell

The experimental CMH corner cell (`general - rmeans - cmeans + cor`, see `CMH-2x2.md`) is an
inclusion-exclusion combination of four different quadratic-form statistics -- algebraically
consistent, but not guaranteed non-negative, and its chi-square distribution under the null is
unverified. `rowcol` vs. the saturated model gives a proper **nested** LR test instead:

```{r residual}
anova(rowcol, saturated, test = "Chisq")
```

Same df as the CMH corner (`(R-2)(C-2) = 8` here) by construction, but a genuine deviance
difference between nested models -- so it's guaranteed >= 0, unlike the CMH corner arithmetic.
The numeric value (`r round(anova(rowcol, saturated)$Deviance[2], 3)`) differs from the CMH
corner's value computed in `CMH-2x2.md` (2.150) -- these are different statistics that happen to
share the same df, not the same quantity computed two ways.

**Takeaway**: if the `print.CMHtest(layout = "2x2")` corner cell proves unreliable, this
`rowcol`-vs-`saturated` residual test is a solid, principled fallback for that specific cell --
at the cost of needing to fit loglinear models rather than just reorganizing `CMHtest()`'s
existing output.
