---
title: "Mosaic plots"
author: "Michael Friendly"
date: "`r Sys.Date()`"
package: vcdExtra
output: 
  rmarkdown::html_vignette:
  fig_caption: yes
bibliography: ["vcd.bib", "vcdExtra.bib"]
csl: apa.csl
vignette: >
  %\VignetteIndexEntry{Mosaic plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  fig.height = 6,
  fig.width = 7,
  fig.path = "fig/tut04-",
  dev = "png",
  comment = "##"
)

# save some typing
knitr::set_alias(w = "fig.width",
                 h = "fig.height",
                 cap = "fig.cap")


# preload datasets ???
set.seed(1071)
library(vcd)
library(vcdExtra)
library(ggplot2)
data(HairEyeColor)
data(PreSex)
data(Arthritis, package="vcd")
art <- xtabs(~Treatment + Improved, data = Arthritis)
if(!file.exists("fig")) dir.create("fig")

```

Mosaic plots provide an ideal method both for visualizing contingency tables and for 
visualizing the fit--- or more importantly--- lack of fit of a loglinear model.
For a two-way table, `mosaic()` fits a model of independence, $[A][B]$
or `~A+B` as an R formula.
For $n$-way tables, `mosaic()` can fit any loglinear model, and can also be
used to plot a model fit with `loglm()`.
See @vcd:Friendly:1994,vcd:Friendly:1999 for the statistical ideas behind these
uses of mosaic displays in connection with loglinear models.

The essential idea is to recursively sub-divide a unit square into rectangular "tiles" for the
cells of the table, such
that the are area of each tile is proportional to the cell frequency.
For a given loglinear model, the tiles can then be shaded in various ways to reflect
the residuals (lack of fit) for a given model.  The pattern of residuals can then
be used to suggest a better model or understand *where* a given model fits or
does not fit.

`mosaic()` provides a wide range of options for the directions of splitting,
the specification of shading, labeling, spacing, legend and many other details.
It is actually implemented as a special case of a more general
class of displays for $n$-way tables called `strucplot`, including
sieve diagrams, association plots, double-decker plots as well as mosaic
plots.  For details, see `help(strucplot)` and the "See also" links,
and also @vcd:Meyer+Zeileis+Hornik:2006b, which is available as
an R vignette via `vignette("strucplot", package="vcd")`.

***Example***:
A mosaic plot for the Arthritis treatment data fits the model of independence,
`~ Treatment + Improved` and displays the association in the pattern of
residual shading.
The plot below is produced with the following call to `mosaic()`.


```{r}
#| Arthritis1,
#| fig.height = 6,
#| fig.width = 7,
#| fig.cap = "Mosaic plot for the `Arthritis` data."
data(Arthritis, package="vcd")
art <- xtabs(~Treatment + Improved, data = Arthritis)
mosaic(art, gp = shading_max, 
            split_vertical = TRUE, 
            main="Arthritis: [Treatment] [Improved]")
```

`gp = shading_max` specifies that color in the plot signals a significant residual at a 90% or 99% significance level,
with the more intense shade for 99%.
Note that the residuals for the independence model were not large
(as shown in the legend),
yet the association between `Treatment` and `Improved`
is highly significant.
```{r, art1}
summary(art)
```

In contrast, one of the other shading schemes, from @vcd:Friendly:1994
(use: `gp = shading_Friendly`), 
uses fixed cutoffs of $\pm 2, \pm 4$,
to shade cells which are *individually* significant
at approximately $\alpha = 0.05$ and $\alpha = 0.001$ levels, respectively.
The right panel below uses `gp = shading_Friendly`.

```{r Arthritis2, h=6, w=7}
mosaic(art, gp = shading_Friendly, 
            split_vertical = TRUE, 
            main="Arthritis: gp = shading_Friendly")
```
