---
title: "1-Creating and manipulating frequency tables"
author: "Michael Friendly"
date: "`r Sys.Date()`"
package: vcdExtra
output: 
  rmarkdown::html_vignette:
  fig_caption: yes
bibliography: ["vcd.bib", "vcdExtra.bib"]
csl: apa.csl
vignette: >
  %\VignetteIndexEntry{Creating and manipulating frequency tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  fig.height = 6,
  fig.width = 7,
  fig.path = "fig/tut01-",
  dev = "png",
  comment = "##"
)

# save some typing
knitr::set_alias(w = "fig.width",
                 h = "fig.height",
                 cap = "fig.cap")

# Old Sweave options
# \SweaveOpts{engine=R,eps=TRUE,height=6,width=7,results=hide,fig=FALSE,echo=TRUE}
# \SweaveOpts{engine=R,height=6,width=7,results=hide,fig=FALSE,echo=TRUE}
# \SweaveOpts{prefix.string=fig/vcd-tut,eps=FALSE}
# \SweaveOpts{keep.source=TRUE}


# preload datasets ???
set.seed(1071)
library(vcd)
library(vcdExtra)
library(ggplot2)
data(HairEyeColor)
data(PreSex)
data(Arthritis, package="vcd")
art <- xtabs(~Treatment + Improved, data = Arthritis)
if(!file.exists("fig")) dir.create("fig")

```


R provides many methods for creating frequency and contingency tables. Several are
described below. In the examples below, we use some real examples and some anonymous
ones, where the variables `A`, `B`, and `C`  represent
categorical variables, and `X` represents an arbitrary R data object.

## Forms of frequency data
The first thing you need to know is that categorical data can be represented in
three different forms in R, and it is sometimes necessary to convert
from one form to another, for carrying out statistical tests, fitting models
or visualizing the results.  Once a data object exists in R, 
you can examine its complete structure with
the `str()` function, or view the names of its components with the
`names()` function.
 

### Case form
Categorical data in case form are simply data frames containing individual observations, with one or
more factors, used as the classifying variables. In case form, there may also
be numeric covariates. 
The total number of observations is `nrow(X)`, and the number of variables is `ncol(X)`.
  	
***Example***:

The `Arthritis` data is available in case form in the `vcd` package. 
There are two explanatory factors: `Treatment` and `Sex`. `Age`
is a numeric covariate, and `Improved` is the response--- an ordered factor,
with levels
`r paste(levels(Arthritis$Improved),collapse=' < ')`.
Excluding `Age`, this represents
a $2 \times 2 \times 3$ contingency table for `Treatment`, `Sex` and `Improved`, but in case form.

```{r, case-form}
names(Arthritis)      # show the variables

str(Arthritis)        # show the structure

head(Arthritis,5)     # first 5 observations, same as Arthritis[1:5,] 
```
  
### Frequency form
Data in frequency form is also a data frame containing one or more factors, and a frequency 
variable, often called `Freq` or `count`.  The total number of observations
is: `sum(X$Freq)`,  `sum(X[,"Freq"])` or some equivalent form.

The number of cells in the table is given by `nrow(X)`.

***Example***: 
For small frequency tables, it is often convenient to enter them in frequency form
using `expand.grid()` for the factors and `c()` to list the counts in a vector.
The example below, from [@vcd:Agresti:2002] gives results for the 1991 General Social Survey,
with respondents classified by sex and party identification.
```{r, frequency-form}
# Agresti (2002), table 3.11, p. 106
GSS <- data.frame(
  expand.grid(sex = c("female", "male"), 
              party = c("dem", "indep", "rep")),
  count = c(279,165,73,47,225,191))

GSS
names(GSS)
str(GSS)

sum(GSS$count)
```

### Table form 
Table form data is represented by a `matrix`, `array` or `table` object, whose elements are the frequencies
in an $n$-way table.  The variable names (factors) and their levels are given by
`dimnames(X)`. The total number of observations
is `sum(X)`.  The number of dimensions of the table is `length(dimnames(X))`,
and the table sizes are given by `sapply(dimnames(X), length)`.
  	
***Example***: 
The `HairEyeColor` is stored in table form in `vcd`.  
```{r, table-form1}
str(HairEyeColor)                      # show the structure

sum(HairEyeColor)                      # number of cases

sapply(dimnames(HairEyeColor), length) # table dimension sizes
```
***Example***: 
Enter frequencies in a matrix, and assign `dimnames`,
giving the variable names and category labels.  Note that, by default,
`matrix()` uses the elements supplied by *columns* in the
result, unless you specify `byrow=TRUE`.
```{r, table-form2}
# A 4 x 4 table  Agresti (2002, Table 2.8, p. 57) Job Satisfaction
JobSat <- matrix(c( 1, 2, 1, 0, 
                    3, 3, 6, 1, 
                   10,10,14, 9, 
                    6, 7,12,11), 4, 4)

dimnames(JobSat) = list(
  income = c("< 15k", "15-25k", "25-40k", "> 40k"),
  satisfaction = c("VeryD", "LittleD", "ModerateS", "VeryS")
  )

JobSat
```

`JobSat` is a **matrix**, not an object of `class("table")`, and some functions
are happier with tables than matrices.
You can coerce it to a table with `as.table()`,
```{r, table-form3}
JobSat <- as.table(JobSat)
str(JobSat)
```

## Ordered factors and reordered tables {#sec:ordered-factors}

In table form, the values of the table factors are ordered by their position in the table.
Thus in the `JobSat` data, both `income` and `satisfaction` represent ordered
factors, and the *positions* of the values in the rows and columns reflects their
ordered nature.

Yet, for analysis, there are time when you need *numeric* values for the levels
of ordered factors in a table, e.g., to treat a factor as a quantitative variable.
In such cases, you can simply re-assign the `dimnames` attribute of the table
variables.  For example, here, we assign numeric values to `income` as the middle of their
ranges, and treat `satisfaction` as equally spaced with integer scores.

```{r, relevel, eval=FALSE}
dimnames(JobSat)$income <- c(7.5,20,32.5,60)
dimnames(JobSat)$satisfaction <- 1:4
```

For the  `HairEyeColor` data, hair color and eye color are ordered arbitrarily.
For visualizing the data using mosaic plots and other methods described below, it 
turns out to be more useful to assure that both hair color and eye color are
ordered from dark to light.
Hair colors are actually ordered this way already, and it is easiest to re-order
eye colors by indexing. Again `str()` is your friend.

```{r, reorder1}
HairEyeColor <- HairEyeColor[, c(1,3,4,2), ]
str(HairEyeColor)
```
This is also the order for both hair color and eye color shown in 
the result of a correspondence analysis (\figref{fig:ca-haireye}) below.

With data in  case form or frequency form, when you have ordered factors
represented with character values, you must ensure that they are treated
as ordered in R.
<!-- \footnote{In SAS, many procedures offer the option -->
<!-- 	`order = data | internal | formatted` to allow character values -->
<!-- 	to be ordered according to (a) their order in the data set, (b) -->
<!-- 	sorted internal value, or (c) sorted formatted representation -->
<!-- 	provided by a SAS format. -->
<!-- } -->

Imagine that the `Arthritis` data was read from a text file.  
By default the `Improved` will be ordered alphabetically:
`Marked`,
`None`,
`Some` --- not what we want.  In this case, the function
`ordered()` (and others) can be useful. 

```{r, reorder2, echo=TRUE, eval=FALSE}
Arthritis <- read.csv("arthritis.txt",header=TRUE)
Arthritis$Improved <- ordered(Arthritis$Improved, 
                              levels=c("None", "Some", "Marked")
                              )
```
The dataset `Arthritis` in the `vcd` package is a data.frame in this form
With this order of  `Improved`, the response in this data,
a mosaic display of `Treatment` and `Improved` (\figref{fig:arthritis}) shows a clearly
interpretable pattern.

The original version of `mosaic` in the `vcd` package required the input to be 
a contingency table in array form, so we convert using `xtabs()`.

<!-- ```{r Arthritis, height=6, width=7, fig.cap="Mosaic plot for the `Arthritis` data ..."} -->

```{r}
#| Arthritis,
#| fig.height = 6,
#| fig.width = 6,
#| fig.cap = "Mosaic plot for the `Arthritis` data, showing the marginal model of independence for Treatment and Improved.  Age, a covariate, and Sex are ignored here."
data(Arthritis, package="vcd")
art <- xtabs(~Treatment + Improved, data = Arthritis)
mosaic(art, gp = shading_max, split_vertical = TRUE, main="Arthritis: [Treatment] [Improved]")
```


Finally, there are situations where, particularly for display purposes, you
want to re-order the *dimensions* of an $n$-way table, or change the
labels for the variables or levels.
This is easy when the data are in table form: `aperm()` permutes
the dimensions, and assigning to `names` and `dimnames`
changes variable names and level labels respectively.
We will use the following version of `UCBAdmissions` in
\@(sec:mantel) below.
^[Changing `Admit` to `Admit?` might be useful for display purposes, but is
dangerous--- because it is then difficult to use that variable name in a model formula.
See \@(sec:tips) for options `labeling_args` and `set_labels`to change variable and level names for displays in the `strucplot` framework.]

```{r, reorder3}
UCB <- aperm(UCBAdmissions, c(2, 1, 3))
dimnames(UCB)[[2]] <- c("Yes", "No")
names(dimnames(UCB)) <- c("Sex", "Admit?", "Department")

# display as a flattened table
stats::ftable(UCB)
```

## `structable()` {#sec:structable}

For 3-way and larger tables
the `structable()` function in `vcd` provides a convenient and flexible tabular display.
The variables assigned to the rows and columns of a two-way display can be specified
by a model formula.
```{r, structable}
structable(HairEyeColor)                   # show the table: default

structable(Hair+Sex ~ Eye, HairEyeColor)   # specify col ~ row variables
```
It also returns an object of class `"structable"` which may be plotted with 
`mosaic()` (not shown here).
```{r, structable1,eval=FALSE}
HSE < - structable(Hair+Sex ~ Eye, HairEyeColor)   # save structable object
mosaic(HSE)                                        # plot it
```

## `table()` and friends {#sec:table}

You  can  generate frequency  tables from factor variables  using the  `table()` function,  tables  of
proportions using  the `prop.table()` function,  and marginal  frequencies using
`margin.table()`.

For these examples, create some categorical vectors:
```{r, table-setup}
 n=500
 A <- factor(sample(c("a1","a2"), n, rep=TRUE))
 B <- factor(sample(c("b1","b2"), n, rep=TRUE))
 C <- factor(sample(c("c1","c2"), n, rep=TRUE))
 mydata <- data.frame(A,B,C)
```

These lines illustrate `table`-related functions:
```{r, table-ex1}
# 2-Way Frequency Table
attach(mydata)
mytable <- table(A,B)   # A will be rows, B will be columns
mytable                 # print table

margin.table(mytable, 1) # A frequencies (summed over B)
margin.table(mytable, 2) # B frequencies (summed over A)

prop.table(mytable)    # cell percentages
prop.table(mytable, 1) # row percentages
prop.table(mytable, 2) # column percentages
```

`table()` can  also  generate  multidimensional  tables  based  on  3  or  more
categorical variables. In  this case, you can use  the `ftable()`  or `structable()`
function to print the results more attractively.

```{r, table-ex2}
# 3-Way Frequency Table
mytable <- table(A, B, C)
ftable(mytable)
```

`table()`  ignores missing values by default. 
To include `NA` as a category in counts, include the
table option  `exclude=NULL` if  the variable  is a  vector. If  the variable is a
factor you  have to  create a  new factor  using \code{newfactor  <- factor(oldfactor,
exclude=NULL)}. 

## `xtabs()` {#sec:xtabs}

The `xtabs()` function allows you to create cross-tabulations of data using formula style input.
This typically works with case-form data supplied in a data frame or a matrix.
The result is a contingency table in array format, whose dimensions are determined by
the terms on the right side of the formula.

```{r, xtabs-ex1}
# 3-Way Frequency Table
mytable <- xtabs(~A+B+C, data=mydata)

ftable(mytable)    # print table

summary(mytable)   # chi-square test of indepedence
```

If a variable is included on the left side of the formula, it is assumed to be a
vector  of  frequencies  (useful if the data have already been tabulated in frequency form).

```{r, xtabs-ex2}
(GSStab <- xtabs(count ~ sex + party, data=GSS))

summary(GSStab)
```

## Collapsing over table factors: `aggregate()`, `margin.table()` and `apply()`}

It sometimes happens that we have a data set with more variables or factors than 
we want to analyse, or else, having done some initial analyses, we decide that
certain factors are not important, and so should be excluded from graphic displays
by collapsing (summing) over them.  For example, mosaic plots and fourfold displays
are often simpler to construct from versions of the data collapsed over
the factors which are not shown in the plots.

The appropriate tools to use again depend on
the form in which the data are represented--- a case-form data frame, a
frequency-form data frame (`aggregate()`), or a table-form array or 
table object (`margin.table()` or `apply()`).

When the data are in frequency form, and we want to produce another
frequency data frame, `aggregate()` is a handy tool, using
the argument `FUN=sum` to sum the frequency variable over the 
factors *not* mentioned in the formula.

***Example***: 
The data frame `DaytonSurvey` in the `vcdExtra` package represents a
$2^5$ table giving the frequencies of reported use (``ever used?'') of 
alcohol, cigarettes and marijuana in a sample of high school seniors,
also classified by sex and race.  

```{r, dayton1}
str(DaytonSurvey)
head(DaytonSurvey)
```

To focus on the associations among the
substances, we want to collapse over sex and race. The right-hand side of the formula
used in the call to `aggregate()` gives the factors to be retained in the
new frequency data frame, `Dayton.ACM.df`.

```{r, dayton2}
# data in frequency form
# collapse over sex and race
Dayton.ACM.df <- aggregate(Freq ~ cigarette+alcohol+marijuana, 
                           data=DaytonSurvey, 
                           FUN=sum)
Dayton.ACM.df
```

When the data are in table form, and we want to produce another
table, `apply()` with `FUN=sum` can be used in a similar way
to sum the table over dimensions not mentioned in the `MARGIN`
argument.  `margin.table()` is just a wrapper for `apply()`
using the `sum()` function.


***Example***: 
To illustrate, we first convert the `DaytonSurvey` to a 5-way
table using `xtabs()`, giving `Dayton.tab`.  

```{r, dayton3}
# in table form
Dayton.tab <- xtabs(Freq ~ cigarette+alcohol+marijuana+sex+race, 
                    data=DaytonSurvey)
structable(cigarette+alcohol+marijuana ~ sex+race, 
           data=Dayton.tab)
```

Then, use `apply()` on `Dayton.tab` to give the
3-way table `Dayton.ACM.tab` summed over sex and race.
The elements in this new table are the column sums for 
`Dayton.tab` shown by `structable()` just above.

```{r, dayton4}
# collapse over sex and race
Dayton.ACM.tab <- apply(Dayton.tab, MARGIN=1:3, FUN=sum)
Dayton.ACM.tab <- margin.table(Dayton.tab, 1:3)   # same result

structable(cigarette+alcohol ~ marijuana, data=Dayton.ACM.tab)
```


Many of these operations can be performed using the `**ply()` functions
in the `plyr` package.
For example, with the data in a frequency form data frame, use `ddply()`
to collapse over unmentioned factors, and `plyr::summarise()`
as the function to be applied to each piece.
<!-- \footnote{ -->
<!-- Ugh. This `plyr` function clashes with a function of the same name in `vcdExtra`. -->
<!-- In this document I will use the explicit double-colon notation to keep them -->
<!-- separate. -->
<!-- } -->

```{r, dayton5, eval=FALSE}
library(plyr)
Dayton.ACM.df <- plyr::ddply(DaytonSurvey, 
                             .(cigarette, alcohol, marijuana), 
                             plyr::summarise, Freq=sum(Freq))

Dayton.ACM.df
```

## Collapsing table levels: `collapse.table()`

A related problem arises when we have a table or array and for some purpose
we want to reduce the number of levels of some factors by summing subsets
of the frequencies.  For example, we may have initially coded Age in 10-year
intervals, and decide that, either for analysis or display purposes, we
want to reduce Age to 20-year intervals.  The `collapse.table()` function
in `vcdExtra` was designed for this purpose.

***Example***:
Create a 3-way table, and collapse Age from 10-year to 20-year intervals. 
First, we generate a $2 \times 6 \times 3$ table of random counts from a 
Poisson distribution with mean of 100.
```{r, collapse1}
# create some sample data in frequency form
sex <- c("Male", "Female")
age <- c("10-19", "20-29",  "30-39", "40-49", "50-59", "60-69")
education <- c("low", 'med', 'high')
data <- expand.grid(sex=sex, age=age, education=education)
counts <- rpois(36, 100)   # random Possion cell frequencies
data <- cbind(data, counts)

# make it into a 3-way table
t1 <- xtabs(counts ~ sex + age + education, data=data)
structable(t1)
``` 
Now collapse `age` to 20-year intervals, and `education`
to 2 levels. In the arguments, levels of `age` and `education`
given the same label are summed in the resulting smaller table.
```{r, collapse2}
# collapse age to 3 levels, education to 2 levels
t2 <- collapse.table(t1, 
         age=c("10-29", "10-29",  "30-49", "30-49", "50-69", "50-69"),
         education=c("<high", "<high", "high"))
structable(t2)
```

## Converting among frequency tables and data frames

As we've seen, a given contingency table can be represented 
equivalently in different forms,
but some R functions were designed for one particular representation.

The table below shows some handy tools for converting from one form to another.

<!-- [htb] -->
<!--  \caption{Tools for converting among different forms for categorical data} {#tab:convert} -->
 
<!--    {llll} -->
<!-- 	\hline -->
<!--                  & \multicolumn{3}{c}{**To this**} \\ -->
<!-- 	**From this**      &     Case form         & Frequency form             &  Table form \\ -->
<!-- 	\hline -->
<!-- 	Case form      &   noop                 & `xtabs(~A+B)`        &  `table(A,B)`  \\  -->
<!-- 	Frequency form &  `expand.dft(X)`  & noop                      & `xtabs(count~A+B)`\\ -->
<!-- 	Table form     &  `expand.dft(X)`  & `as.data.frame(X)`   &  noop \\ -->
<!-- 	\hline -->
   
| **From this**    |                     | **To this**          |                   |
|:-----------------|:--------------------|:---------------------|-------------------|
|	                 |    _Case form_      | _Frequency form_     | _Table form_      |
|	_Case form_      |   noop              | `xtabs(~A+B)`        |  `table(A,B)`     |
|	_Frequency form_ |  `expand.dft(X)`    | noop                 | `xtabs(count~A+B)`|
|	_Table form_     |  `expand.dft(X)`    | `as.data.frame(X)`   |  noop             |


For example, a contingency table in table form (an object of `class(table)`) can be converted
to a data.frame with `as.data.frame()`.
^[Because R is object-oriented, this is actually a short-hand for the function `as.data.frame.table()`.] 
The resulting
`data.frame` contains columns
representing the classifying factors and the table entries (as a column named by
the `responseName` argument, defaulting to `Freq`. This is the inverse of `xtabs()`.

***Example***: 
Convert the `GSStab` in table form to a data.frame in frequency form.
```{r, convert-ex1}
as.data.frame(GSStab)
```

***Example***: Convert the `Arthritis` data in case form to a 3-way table of
`Treatment` $\times$ `Sex` $\times$ `Improved`.
Note the use of `with()` to avoid having to use `Arthritis\$Treatment` etc. within the call to `table()`.%
^[`table()` does not allow a `data` argument to provide an environment in which the table variables are to be found.  In the examples in \@(sec:table) I used `attach(mydata)` for this purpose,
but `attach()` leaves the variables in the global environment, while `with()` just evaluates the `table()` expression in a temporary environment of the data.]

```{r, convert-ex2}
Art.tab <- with(Arthritis, table(Treatment, Sex, Improved))
str(Art.tab)

ftable(Art.tab)
```

There may also be times that you will need an equivalent case form `data.frame`
with factors  representing the table variables
rather than the frequency  table.
For example, the `mca()` function in package `MASS`
only operates on data in this format. 
Marc Schwartz initially provided code for `expand.dft()` on the  Rhelp
mailing list for converting a table back into a case form `data.frame`.
This function is included in `vcdExtra`.

***Example***: Convert the `Arthritis` data in table form (`Art.tab`) back to a `data.frame`
in case form, with factors
`Treatment`, `Sex` and `Improved`.
```{r, convert-ex3}
Art.df <- expand.dft(Art.tab)
str(Art.df)
```

## A complex example {#sec:complex}

If you've followed so far, you're ready for a more complicated example.
The data file, `tv.dat` represents a 4-way table of size
$5 \times 11 \times 5 \times 3$ where the table variables (unnamed in the file)
are read as `V1` -- `V4`, and the cell frequency is read
as `V5`.  The file, stored in the `doc/extdata` directory
of `vcdExtra`, can be read as follows:
```{r, tv1}
tv.data<-read.table(system.file("extdata","tv.dat", package="vcdExtra"))
head(tv.data,5)
```
For a local file, just use `read.table()` in this form:
```{r, tv2,eval=FALSE}
tv.data<-read.table("C:/R/data/tv.dat")
```
The data `tv.dat` came from the initial implementation of 
mosaic displays in R by Jay Emerson. 
In turn, they came from the initial development of mosaic displays 
[@vcd:Hartigan+Kleiner:1984]
that illustrated the method with data on a large sample of TV viewers
whose behavior had been recorded for the Neilson ratings.
This data set contains sample television audience data from Neilsen
Media Research for the week starting November 6, 1995.

The table variables are:

* `V1`-- values 1:5 correspond to the days Monday--Friday;
* `V2`-- values 1:11 correspond to the quarter hour times 8:00PM through 10:30PM;
* `V3`-- values 1:5 correspond to ABC, CBS, NBC, Fox, and non-network choices;
* `V4`-- values 1:3 correspond to transition states: turn the television Off, Switch channels, or Persist in viewing the current channel.

We are interested just the cell frequencies, and rely on the facts that the

(a) the table is complete--- there are no missing cells, so `nrow(tv.data)` = `r nrow(tv.data)`;
(b) the observations are ordered so that `V1` varies most rapidly and `V4` most slowly.  From this, we can just extract the frequency column and reshape it into an array. [That would be dangerous if any observations were out of order.]

```{r, tv3}
TV <- array(tv.data[,5], dim=c(5,11,5,3))                                        
dimnames(TV) <- list(c("Monday","Tuesday","Wednesday","Thursday","Friday"), 
                     c("8:00","8:15","8:30","8:45","9:00","9:15","9:30",         
                       "9:45","10:00","10:15","10:30"),                            
                     c("ABC","CBS","NBC","Fox","Other"), 
                     c("Off","Switch","Persist"))

names(dimnames(TV))<-c("Day", "Time", "Network", "State")
```

More generally (even if there are missing cells), we can 
use `xtabs()` (or `plyr::daply()`)
to do the cross-tabulation, using `V5` as the
frequency variable.  Here's how to do this same operation with `xtabs()`:
```{r, tv3a,eval=FALSE}
TV <- xtabs(V5 ~ ., data=tv.data)
dimnames(TV) <- list(Day = c("Monday","Tuesday","Wednesday","Thursday","Friday"), 
                     Time = c("8:00","8:15","8:30","8:45","9:00","9:15","9:30",         
                              "9:45","10:00","10:15","10:30"),                            
                     Network = c("ABC","CBS","NBC","Fox","Other"), 
                     State = c("Off","Switch","Persist"))

# table dimensions
dim(TV)
```

But this 4-way table is too large and awkward to work with. Among the networks,
Fox and Other occur infrequently. 
We can also cut it down to a 3-way table by considering only viewers who persist
with the current station.
^[This relies on the fact that that indexing an array drops dimensions of length 1 by default, using the argument `drop=TRUE`; the result is coerced to the lowest possible dimension.]


```{r, tv4}
TV2 <- TV[,,1:3,]      # keep only ABC, CBS, NBC
TV2 <- TV2[,,,3]       # keep only Persist -- now a 3 way table
structable(TV2)
```

Finally, for some purposes, we might want to collapse the 11 times into a smaller number.
Half-hour time slots make more sense.
Here, we use `as.data.frame.table()` to convert the table back to a data frame,
 `levels()` to re-assign the values of `Time`,
and finally, `xtabs()` to give a new, collapsed frequency table.

```{r, tv5}
TV.df <- as.data.frame.table(TV2)
levels(TV.df$Time) <- c(rep("8:00", 2),
                        rep("8:30", 2),
                        rep("9:00", 2), 
                        rep("9:30", 2), 
                        rep("10:00",2),
                            "10:30"
                        )

TV3 <- xtabs(Freq ~ Day + Time + Network, TV.df)

structable(Day ~ Time+Network, TV3)
```

We've come this far, so we might as well show a mosaic display. This is analogous to that used by
@vcd:Hartigan+Kleiner:1984. 
```{r tv-mosaic1, fig.height=6, fig.width=6}
mosaic(TV3, shade = TRUE,
       labeling = labeling_border(rot_labels = c(0, 0, 0, 90)))
```

This mosaic displays can be read at several levels, corresponding to the successive splits of
the tiles and the residual shading.
Several trends are clear for viewers who persist:

* Overall, there are about the same number of viewers on each weekday, with slightly more on Thursday.
* Looking at time slots, viewership is slightly greater from 9:00 - 10:00 overall and also
8:00 - 9:00 on Thursday and Friday

From the residual shading of the tiles:

* Monday: CBS dominates in all time slots.
* Tuesday" ABC and CBS  dominate after 9:00
* Thursday: is a largely NBC day
* Friday: ABC dominates in the early evening

<!-- ```{r, tv4} -->
<!-- TV.df <- as.data.frame.table(TV) -->
<!-- levels(TV.df$Time) <- c(rep("8:00-8:59",4), -->
<!--                         rep("9:00-9:59",4),  -->
<!--                         rep("10:00-10:44",3)) -->
<!-- TV3 <- xtabs(Freq ~ Day + Time + Network, TV.df) -->

<!-- structable(Day ~ Time+Network, TV3) -->
<!-- ``` -->


<!-- Whew! See \figref{fig:TV-mosaic} for a mosaic plot of the `TV3` data. -->

<!-- The table is too large to display conveniently, but we can show a subtable -->
<!-- by selecting the indices.  This call to `ftable()` subsets the first three -->
<!-- levels of each factor. -->

<!-- ```{r tv4} -->
<!-- ftable(TV[1:3,1:3,1:3,1:3], col.vars = 3:4) -->
<!-- ``` -->

<!-- We've come this far, so we might as well show a mosaic display. This is analogous to that used by -->
<!-- @vcd:Hartigan+Kleiner:1984. The result is too complex to see very much.  It would be useful to simplify -->
<!-- the table by collapsing one or more of the table dimensions, e.g., `Time`. -->

<!-- ```{r tv-mosaic2, fig.height=7, fig.width=7} -->
<!-- mosaic(TV[,,1:4,], shade=TRUE,  -->
<!--        labeling_args = list(abbreviate_labs = c(8,5,1,1))) -->

<!-- ``` -->



<!-- ## Styles -->

<!-- The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows: -->

<!--     output:  -->
<!--       rmarkdown::html_vignette: -->
<!--         css: mystyles.css -->

<!-- ## Figures -->

<!-- The figure sizes have been customised so that you can easily put two images side-by-side.  -->

<!-- ```{r, fig.show='hold'} -->
<!-- plot(1:10) -->
<!-- plot(10:1) -->
<!-- ``` -->


# References

