---
title: "Tests of Independence"
author: "Michael Friendly"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
  fig_caption: yes
bibliography: ["vcd.bib", "vcdExtra.bib"]
vignette: >
  %\VignetteIndexEntry{Tests of Independence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  fig.height = 6,
  fig.width = 7,
  fig.path = "fig/tut02-",
  dev = "png",
  comment = "##"
)

# save some typing
knitr::set_alias(w = "fig.width",
                 h = "fig.height",
                 cap = "fig.cap")

# Old Sweave options
# \SweaveOpts{engine=R,eps=TRUE,height=6,width=7,results=hide,fig=FALSE,echo=TRUE}
# \SweaveOpts{engine=R,height=6,width=7,results=hide,fig=FALSE,echo=TRUE}
# \SweaveOpts{prefix.string=fig/vcd-tut,eps=FALSE}
# \SweaveOpts{keep.source=TRUE}


# preload datasets ???
set.seed(1071)
library(vcd)
library(vcdExtra)
library(ggplot2)
data(HairEyeColor)
data(PreSex)
data(Arthritis, package="vcd")
art <- xtabs(~Treatment + Improved, data = Arthritis)
if(!file.exists("fig")) dir.create("fig")

```

OK, now we're ready to do some analyses.  This vignette focuses on relatively simple non-parametric
tests and measures of association.

## CrossTable

For tabular displays,
the  `CrossTable()`  function in  the `gmodels`  package produces  cross-tabulations
modeled after `PROC FREQ` in SAS or `CROSSTABS` in SPSS. 
It has a wealth of options for the quantities that can be shown in each cell.

Recall the GSS data used earlier.
```{r, GSStab}
# Agresti (2002), table 3.11, p. 106
GSS <- data.frame(
  expand.grid(sex = c("female", "male"), 
              party = c("dem", "indep", "rep")),
  count = c(279,165,73,47,225,191))

(GSStab <- xtabs(count ~ sex + party, data=GSS))
```

Generate a crosstable showing cell frequency and the cell contribution to $\chi^2$.
```{r, xtabs-ex2}
# 2-Way Cross Tabulation
library(gmodels)
CrossTable(GSStab, prop.t=FALSE, prop.r=FALSE, prop.c=FALSE)
```

There are  options to  report percentages  (row, column,  cell), specify decimal
places, produce Chi-square,  Fisher, and McNemar  tests of independence,  report
expected  and residual  values (pearson,  standardized, adjusted  standardized),
include missing values as valid, annotate with row and column titles, and format
as SAS or SPSS style output! See `help(CrossTable)` for details.

## Chi-square test

For 2-way tables you can use `chisq.test()` to test independence of the row
and column variable. By default,  the $p$-value is calculated from  the asymptotic
chi-squared distribution of the test  statistic. Optionally, the $p$-value can  be
derived via Monte Carlo simulation. 

```{r, chisq}
(HairEye <- margin.table(HairEyeColor, c(1, 2)))

chisq.test(HairEye)

chisq.test(HairEye, simulate.p.value = TRUE)
```

## Fisher Exact Test {#sec:Fisher}

`fisher.test(X)` provides an  **exact test** of  independence. `X` must be  a two-way
contingency table in table form.  Another form, 
`fisher.test(X, Y)` takes two
categorical vectors of the same length.  
For tables larger than $2 \times 2$ the method can be computationally intensive (or can fail) if
the frequencies are not small.

```{r fisher}
fisher.test(GSStab)
```

Fisher's test is meant for tables with small total sample size.
It generates an error for the `HairEye` data with $n$=592 total frequency.

```{r fisher-error, error=TRUE}
fisher.test(HairEye)
```
## Mantel-Haenszel test and conditional association {#sec:mantel}

Use the  `mantelhaen.test(X)` function  to perform  a Cochran-Mantel-Haenszel 
$\chi^2$ chi
test  of   the  null  hypothesis   that  two  nominal   variables  are
*conditionally independent*, $A \perp B \; | \; C$, in each stratum,  assuming that there is no  three-way
interaction. `X` is  a 3 dimensional  contingency table, where  the last dimension
refers to the strata.

The `UCBAdmissions` serves as an example of a $2 \times 2 \times 6$ table,
with `Dept` as the stratifying variable.
```{r, mantel1}
# UC Berkeley Student Admissions
mantelhaen.test(UCBAdmissions)
```

The results show no evidence for association between admission and gender
when adjusted for department.  However, we can easily see that the assumption
of equal association across the strata (no 3-way association) is probably
violated. For $2 \times 2 \times k$ tables, this can be examined
from the odds ratios for each $2 \times 2$ table (`oddsratio()`), and
tested by  using `woolf_test()` in `vcd`.

```{r, mantel2}
oddsratio(UCBAdmissions, log=FALSE)

lor <- oddsratio(UCBAdmissions)  # capture log odds ratios
summary(lor)

woolf_test(UCBAdmissions) 
``` 

## Some plot methods

### Fourfold displays

We  can visualize the  odds ratios of  Admission for
each  department  with  fourfold  displays  using  `fourfold()`.  The cell
frequencies $n_{ij}$  of each  $2 \times  2$ table  are shown  as a  quarter circle whose
radius is proportional to $\sqrt{n_{ij}}$, so  that its area is proportional to  the
cell frequency.

```{r, reorder3}
UCB <- aperm(UCBAdmissions, c(2, 1, 3))
dimnames(UCB)[[2]] <- c("Yes", "No")
names(dimnames(UCB)) <- c("Sex", "Admit?", "Department")
```
Confidence rings for the odds ratio allow a visual test of the null of no association; 
the rings for adjacent quadrants overlap *iff* the observed counts are consistent 
with the null hypothesis.  In the extended version (the default), brighter colors
are used where the odds ratio is significantly different from 1.
The following lines produce \figref{fig:fourfold1}.
<!-- \footnote{The color values `col[3:4]` were modified from their default values -->
<!-- to show a greater contrast between significant and insignificant associations here.} -->



```{r}
#| fourfold1, 
#| h=5, w=7.5, 
#| cap = "Fourfold display for the `UCBAdmissions` data. Where the odds ratio differs
#|   	significantly from 1.0, the confidence bands do not overlap, and the circle quadrants are
#|  	shaded more intensely."
col <- c("#99CCFF", "#6699CC", "#F9AFAF", "#6666A0", "#FF0000", "#000080")
fourfold(UCB, mfrow=c(2,3), color=col)
```
 

Another `vcd` function, `cotabplot()`, provides a more general approach
to visualizing conditional associations in contingency tables,
similar to trellis-like plots produced by `coplot()` and lattice graphics.
The `panel` argument supplies a function used to render each conditional 
subtable. The following gives a display (not shown) similar to \figref{fig:fourfold1}.
```{r fourfold2, eval=FALSE}
cotabplot(UCB, panel = cotab_fourfold)
```

### Doubledecker plots

When we want to view the conditional
probabilities of a response variable (e.g., `Admit`)
in relation to several factors,
an alternative visualization is a `doubledecker()` plot.
This plot is a specialized version of a mosaic plot, which
highlights the levels of a response variable (plotted vertically)
in relation to the factors (shown horizontally). The following
call produces \figref{fig:doubledecker}, where we use indexing
on the first factor (`Admit`) to make `Admitted`
the highlighted level.

In this plot, the
association between `Admit` and `Gender` is shown
where the heights of the highlighted conditional probabilities
do not align. The excess of females admitted in Dept A stands out here.

```{r}
#| doubledecker, 
#| h=5, w=8,
#| out.width = "70%",
#| cap = "Doubledecker display for the `UCBAdmissions` data. The heights
#|     of the highlighted bars show the conditional probabilities of `Admit`,
#|     given `Dept` and `Gender`."
doubledecker(Admit ~ Dept + Gender, data=UCBAdmissions[2:1,,])
```

### Odds ratio plots
Finally, the there is a `plot()` method for `oddsratio` objects.
By default, it shows the 95% confidence interval for the log odds ratio.
\figref{fig:oddsratio} is produced by:

```{r}
#| oddsratio,
#| h=6, w=6,
#| out.width = "60%",
#| cap = "Log odds ratio plot for the `UCBAdmissions` data."
plot(lor, 
     xlab="Department", 
     ylab="Log Odds Ratio (Admit | Gender)")
```
 {#fig:oddsratio}

## Cochran-Mantel-Haenszel tests for ordinal factors {#sec:CMH}

The standard $\chi^2$ tests for association in a two-way table
treat both table factors as nominal (unordered) categories.
When one or both factors of a two-way table are
quantitative or ordinal, more powerful tests of association
may be obtaianed by taking ordinality into account, using
row and or column scores to test for linear trends or differences
in row or column means.

More general versions of the CMH tests (Landis etal., 1978) 
[@Landis-etal:1978] are provided by assigning
numeric scores to the row and/or column variables. 
For example, with two ordinal factors (assumed to be equally spaced), assigning
integer scores, `1:R` and `1:C` tests the linear $\times$ linear component
of association. This is statistically equivalent to the Pearson correlation between the
integer-scored table variables, with $\chi^2 = (n-1) r^2$, with only 1 $df$
rather than $(R-1)\times(C-1)$ for the test of general association.

When only one table
variable is ordinal, these general CMH tests are analogous to an ANOVA, testing
whether the row mean scores or column mean scores are equal, again consuming
fewer $df$ than the test of general association.

The `CMHtest()` function in `vcdExtra` calculates these various
CMH tests for two possibly ordered factors, optionally stratified other factor(s).

***Example***:
```{r, table-form2, include=FALSE}
## A 4 x 4 table  Agresti (2002, Table 2.8, p. 57) Job Satisfaction
JobSat <- matrix(c(1,2,1,0, 3,3,6,1, 10,10,14,9, 6,7,12,11), 4, 4)
dimnames(JobSat) = list(income=c("< 15k", "15-25k", "25-40k", "> 40k"),
                satisfaction=c("VeryD", "LittleD", "ModerateS", "VeryS"))
JobSat <- as.table(JobSat)
```

Recall the $4 \times 4$ table, `JobSat` introduced in \secref{sec:creating},
```{r, jobsat}
JobSat
```

Treating the `satisfaction` levels as equally spaced, but using
midpoints of the `income` categories as row scores gives the following results:
```{r, cmh1}
CMHtest(JobSat, rscores=c(7.5,20,32.5,60))
```

Note that with the relatively small cell frequencies, the test for general 
give no evidence for association. However, the the `cor` test for linear x linear
association on 1 df is nearly significant. The `coin` package contains the
functions `cmh_test()` and `lbl_test()`
for CMH tests of general association and linear x linear association respectively.

## Measures of Association

There are a variety of statistical measures of *strength* of association for
contingency tables--- similar in spirit to $r$ or $r^2$ for continuous variables.
With a large sample size, even a small degree of association can show a 
significant $\chi^2$, as in the example below for the `GSS` data.

The  `assocstats()`  function in `vcd`  calculates  the   $\phi$
contingency coefficient,  and Cramer's  V for  an $r \times c$  table. 
The input must be in table form, a two-way $r \times c$ table.  
It won't work with `GSS` in frequency form, but by now you should know how
to convert.
```{r, assoc1}
assocstats(GSStab)
```

For tables with ordinal variables, like `JobSat`, some people prefer the
Goodman-Kruskal $\gamma$ statistic
[@vcd:Agresti:2002, \S 2.4.3]
based on a comparison of concordant
and discordant pairs of observations in the case-form equivalent of a two-way table.
```{r, gamma}
GKgamma(JobSat)
```

A web article by Richard Darlington,
[http://node101.psych.cornell.edu/Darlington/crosstab/TABLE0.HTM]
gives further description of these and other measures of association.

## Measures of Agreement
The
`Kappa()` function in the `vcd` package calculates Cohen's $\kappa$ and weighted
$\kappa$ for a square two-way table with the same row and column categories [@Cohen:60].
\footnote{ 
Don't confuse this with `kappa()` in base R that computes something
entirely different (the condition number of a matrix).
}
Normal-theory $z$-tests are obtained by dividing $\kappa$ by its asymptotic standard
error (ASE).  A `confint()` method for `Kappa` objects provides confidence intervals.
```{r, kappa}
data(SexualFun, package = "vcd")
(K <- Kappa(SexualFun))
confint(K)
```

A visualization of agreement [@Bangdiwala:87], both unweighted and weighted for degree of departure
from exact agreement is provided by the `agreementplot()` function.
\figref{fig:agreesex} shows the agreementplot for the `SexualFun` data,
produced as shown below. 

The Bangdiwala measures (returned by the function)
represent the proportion of the
shaded areas of the diagonal rectangles, using weights $w_1$ for exact agreement,
and $w_2$ for partial agreement one step from the main diagonal.

```{r}
#| agreesex, 
#| h=6, w=7,
#| out.width = "60%",
#| cap = "Agreement plot for the `SexualFun` data."
agree <- agreementplot(SexualFun, main="Is sex fun?")
unlist(agree)
```


## References
