---
title: "Demo - Occupational Status"
author: "Michael Friendly"
date: "`r Sys.Date()`"
package: vcdExtra
output: 
  rmarkdown::html_vignette:
  fig_caption: yes
bibliography: ["vcd.bib", "vcdExtra.bib", "vignettes.bib"]
csl: apa.csl
vignette: >
  %\VignetteIndexEntry{Demo - Occupational Status}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.height = 6,
  fig.width = 7,
  fig.path = "fig/demo-occStatus-",
  dev = "png",
  comment = "##"
)

# save some typing
knitr::set_alias(w = "fig.width",
                 h = "fig.height",
                 cap = "fig.cap")

# colorize text
colorize <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{%s}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", color,
      x)
  } else x
}
```

This vignette was one of a series of `demo()` files in the package. It is still there
as `demo("occStatus")`, but is now presented here with additional commentary and analysis,
designed to highlight some aspects of analysis of categorical data and graphical display.

## Social mobility

Social mobility is an important concept in sociology, and its' study has led to
a wide range of developments in categorical data analysis in what are often called
_mobility tables_. 

The idea is to study the movement of individuals, families, households or other categories of people within or between social strata in a society, across time or space. It refers to
a change in social status relative to one's current social location within a given society.
Using survey data, the most frequent examples relate to changes in income or wealth, but most often this is studied via classification in occupational categories ("professional, "managerial", "skilled manual", ...). Most often this is studied _intergenerationaly_ using the occupational
categories of fathers and sons.

Mobility tables are nearly always _square_ tables, with the same categories for the row
and column variables. As such, they nearly always exhibit positive associations
along the diagonal cells.  What is of interest are specialized models, intermediate
between the null model of independence and the saturated model.


The `vcdExtra` package contains a number of datasets on the topic of "mobility".
`help.search("mobility", fields="concept")` gives the datasets:
`vcdExtra::Glass`,
`vcdExtra::Hauser79`,
`vcdExtra::Mobility`,
`vcdExtra::Yamaguchi87`,
with similar characteristics.

### `occupationalStatus` dataset

Here, I focus on a more classic dataset, `datasets::occupationalStatus`,
a cross-classification of `r sum(occupationalStatus)` British males
according to a person's occupational category and that of his father.

This dataset is often attributed to @Goodman:79 (Table 3), but it derives
from a classic study _Social Mobility in Britain_ by @Glass:54.

This dataset is an 8 x 8 table of `r sum(occupationalStatus)` cases.
`origin` refers to the son's occupational
category; `destination` to the father's. @Glass:54 lists these in order
from highest status (1) to lowest (8), corresponding to 
professional, managerial, upper non-man, lower non-man, ... unskilled.

```{r data}
data(occupationalStatus, package="datasets")
str(occupationalStatus)
occupationalStatus
```

### Load packages

```{r load}
library(vcdExtra)
library(gnm)
```

### Mosaic plots

`occupationalStatus` is a `table` object, and the
simplest plot for the frequencies is the default `plot()` method, giving a `graphics::mosaicplot()`.

```{r mosaicplot}
plot(occupationalStatus, shade=TRUE)
```

The total frequencies in the table are first classified by son's occupational category
(`origin`). It is easy to see that category 6 of son's status is most common and the diagonal
cells, where father's status is the same as son's are much more frequent than other combinations. `mosaicplot()`, using `shade=TRUE` colors the tiles according to the 
sign and magnitude of the residuals from an independence model: shades of blue for
positive residuals and red for negative residuals. 

`vcd::mosaic()` gives a similar display, but is much more flexible in the labeling of
the row and column variable, labels for the categories, and the scheme used for shading
the tiles. Here, I simply assign longer labels for the row and column variables,
using the `labeling_args` argument to `mosaic()`.

```{r}
long.labels <- list(set_varnames = c(origin="origin: Son's status", 
                                     destination="destination: Father's status"))

mosaic(occupationalStatus, shade=TRUE, 
       main="Occupational status: Independence model", 
       labeling_args = long.labels, 
       legend=FALSE)
```

### Fitting and graphing models

The call to `vcd::mosaic()` above takes the `occupationalStatus` table as input.
Internally, it fits the model of independence and displays the result,
but for more complex tables, control of the fitted model is limited.

`vcdExtra::mosaic.glm()` is a `mosaic` method for `glm` objects.  
This means you can fit any model, and supply the model object to `mosaic()`.

```{r indep}
indep<- glm(Freq ~ origin + destination, 
            family = poisson, 
            data=occupationalStatus)

# the same mosaic, using the fitted model
mosaic(indep, 
       main="Independence model",
       labeling_args = long.labels, 
       legend=FALSE, 
       gp=shading_Friendly)

```

### Quasi-independence

Among the most important advances from the social mobility literature is the idea that
associations between row and column variables in square tables can be explored in greater
depth if we ignore the obvious association in the diagonal cells.

The result is a model of _quasi-independence_, asserting that fathers' and sons' occupations
are independent, ignoring the diagonal cells. In the [`gnm`]( https://CRAN.R-project.org/package=gnm) package,
`gnm::Diag()` creates the appropriate term in the model formula

```{r quasi}
quasi <- gnm(Freq ~ origin + destination + Diag(origin, destination), 
             family=poisson, 
             data=occupationalStatus)

```


## References
