---
title: "Bug: `mcaplot()` fails with non-conformable arguments in `cacoord()`"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
---

## Summary

`vcdExtra::mcaplot()` throws `Error in rsc %*% diag(sv) : non-conformable arguments`
for certain `mjca` objects. The error fires inside `ca::cacoord()`, called from
`mcaplot()`. It does not occur with the standard examples in `?mcaplot` (Titanic,
HairEyeColor), which is why it went undetected.

## Reproducible example

The data are a 5 × 3 × 2 contingency table of app users (n = 30,364 after
removing cases with missing Gender) cross-classified by engagement cluster,
Gender, and opioid medication use.

```{r setup}
library(ca)
library(vcdExtra)

load("dat_categor-noNA.rda")   # loads dat_categor.noNA: n=30364, vars: cluster/Gender/Opi/Can/DAC

CTO <- xtabs(~ cluster + Gender + Opi, data = dat_categor.noNA)
CTO.mca <- mjca(CTO)
```

`mcaplot()` fails; `ca::plot()` on the same object works fine:

```{r error, error=TRUE}
mcaplot(CTO.mca)   # Error in rsc %*% diag(sv) : non-conformable arguments
ca::plot(CTO.mca)  # OK
```

## Diagnosis

The error originates in `cacoord()` (ca package), which `mcaplot()` calls as:

```r
coords <- cacoord(obj, type = map, rows = FALSE)
```

With the default `dim = NA`, `cacoord()` extracts coordinates using the full
`colcoord` and `sv` slots of the `mjca` object:

```r
sv  <- obj$sv           # length = nd.max
csc <- obj$colcoord     # ncol   = K0
...
cpc <- csc %*% diag(sv) # fails when ncol(csc) != length(sv)
```

Inspecting the `mjca` object reveals the mismatch:

```{r inspect}
cat("obj$nd           :", CTO.mca$nd,              "\n")  # 2  (user-specified)
cat("length(obj$sv)   :", length(CTO.mca$sv),      "\n")  # 2  (nd.max adjusted eigenvalues)
cat("ncol(obj$colcoord):", ncol(CTO.mca$colcoord), "\n")  # 3  (K0 eigenvectors)
```

`colcoord` has **3 columns** but `sv` has only **2 elements**, so
`diag(sv)` is 2×2 while `csc` is J×3 — non-conformable.

## Root cause: a latent bug in `mjca()` (ca package)

Under the default `lambda = "adjusted"` method, `mjca()` computes coordinates
and singular values from **two different eigendecompositions** of **two different
matrices**, and stores them without ensuring they have consistent dimensions:

```r
# From the full Burt matrix (S):
nd.max <- sum(sqrt(lambda0) >= 1/Q, na.rm = TRUE)   # adjusted eigenvalues >= 1/Q

# From the off-diagonal Burt matrix (S.null, diagonal blocks zeroed out):
K0     <- length(which(evd.S.null$values > 1e-8))   # numerically non-trivial eigenvectors
col.sc <- diag(1/sqrt(cm)) %*% evd.S.null$vectors[, 1:K0]  # → colcoord has K0 cols

# Adjusted singular values (length = nd.max) → sv
lambda.adj <- ((Q/(Q-1))^2 * (sqrt(lambda0)[1:nd.max] - 1/Q)^2)
lambda0    <- lambda.adj
...
list(sv = sqrt(lambda0),   # length nd.max
     colcoord = col.sc,    # ncol   K0      ← mismatch when K0 != nd.max
     nd = nd, ...)
```

Normally `K0 == nd.max` and no error occurs. The mismatch arises here because
the binary variable **Opi** (Yes/No, Q = 3 variables, threshold = 1/Q = 1/3)
contributes one eigenvector that is numerically non-trivial in S.null
(eigenvalue > 1e-8, so it enters `colcoord`), but whose **adjusted** eigenvalue
falls *below* the Greenacre threshold 1/Q = 1/3 (so it is excluded from `sv`).
The result is `K0 = 3`, `nd.max = 2`.

This is the first known example to trigger the latent K0 ≠ nd.max condition,
because previous examples (Titanic, HairEyeColor) happened to produce K0 == nd.max.
The binary variable with a particular eigenvalue structure is the triggering
condition.

## Fix applied in `vcdExtra::mcaplot()`

Pass `dim = seq_len(obj$nd)` to `cacoord()` so that only the first `obj$nd`
columns of `colcoord` are used, matching the length of `obj$sv`:

```r
# nd=obj$nd ensures coordinate dimensions match obj$sv length,
# preventing a non-conformable error in rsc %*% diag(sv) inside cacoord()
# when colcoord has more columns (K0) than sv has elements (nd.max).
coords <- cacoord(obj, type = map, rows = FALSE, dim = seq_len(obj$nd))
```

Note: the underlying inconsistency — `ncol(colcoord) != length(sv)` — is a bug
in `ca::mjca()`. A proper fix there would truncate `col.sc` to `nd.max` columns
before storing it as `colcoord`. The fix above is a workaround on the
`vcdExtra` side.

```{r fixed}
mcaplot(CTO.mca)   # now works after the fix
```
